---
title: "Confidence Intervals"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(assertthat)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.eval = FALSE)
knitr::opts_chunk$set(exercise.checker = gradethis::grade_learnr)

```

## Confidence Intervals

### What is a confidence interval?


```{r ci_def_question}
question("Which of the following is the correct definition for a confidence interval for the mean?",
  answer("mean ± critical value * standard error", correct=TRUE),
  answer("mean ± critical value / standard deviation"),
  answer("mean ± critical value * standard deviation"),
  answer("mean ± mean * standard error"),
  random_answer_order = TRUE
)
```


```{r ci_interp_question}
question("We collect a sample and compute the confidence interval for the mean, and find that it spans from 3.2 to 6.8.  What is the most appropriate interpretation of this finding?",
  answer("If we were to repeat this procedure across multiple samples, the resulting confidence interval for each sample will contain the true population mean 95% of the time", correct=TRUE),
  answer("If we were to repeat this procedure across multiple samples, 95% of the time the true population mean will fall between 3.2 and 6.8."),
  answer("There is a probability of .95 that the true population mean lies between 3.2 and 6.8"),
  answer("We have 95% confidence that the true population mean lies between 3.2 and 6.8"),
  random_answer_order = TRUE
)
```


### Computing a confidence interval using the normal distribution

As we saw in the previous quiz, the definition for a confidence interval for the mean is:

$$
CI = \text{mean} \pm \text{critical value} * \text{standard error}
$$

Let's first create some data to use as an example. We will sample 100 values from a standard normal distribution (i.e. with a mean of zero and a standard devation of one).  

```{r ci_setup, echo=FALSE}
set.seed(12345)
sample_size <- 100
my_sample <- rnorm(100)
critical_value <- qnorm(0.975)
my_sample_mean <- mean(my_sample)
my_sample_sd <- sd(my_sample)

```

```{r sample, exercise=TRUE, exercise.setup='ci_setup'}
sample_size <- 100
my_sample <- rnorm(sample_size)

```

#### Exercise

Compute the sample mean and standard deviation for this sample.

```{r samplemean, exercise=TRUE, exercise.setup='ci_setup'}
my_sample_mean <- ...
my_sample_sd <- ...
```

```{r samplemean-solution}
my_sample_mean <- mean(my_sample)
my_sample_sd <- sd(my_sample)
```

```{r samplemean-check}
grade_code(incorrect='Try again...')

```

### Determining the critical value

The critical value is based on a known distribution -- in this case, the standard normal distribution.  We want to find the values of our distribution that cutoff a particular percentage of the distribution in the tails.  For example, for a 95% confidence interval, we want to find the values that contain 95% of the distribution, leaving 5% in the tails. This means that we need to find the cutoffs that leave 2.5% of the distribution in each tail (upper and lower), so that in total there will be 5% of the distribution outside of our interval.  These are the 2.5th and 97.5th percentiles of the distribution. 

We can compute the value of the normal distribution that cuts off a particular amount of the distribution using the *quantile* function for that distribution; for the normal distribution, this is the `qnorm()` function.  We have created a special function called `plot_normal_quantile()` that allows you to see the results for different quantiles. In the following cell, try entering various values between zero and one to see what the resulting cutoff is.

```{r normal_quantile_plot, echo=FALSE}


plot_normal_quantile <- function(q){
  if (q <= 0 | q >= 1){
    return('ERROR: q must be between zero and one')
  }
  normdf = data.frame(xval=seq(-4, 4, 0.01)) %>%
    mutate(dnorm=dnorm(xval))
  cutoff = qnorm(q)
  ggplot(normdf, aes(xval,dnorm)) + 
    geom_line() + 
    xlab('value') + 
    ylab('density') + 
    ylim(0, 0.45) +
    geom_vline(xintercept=cutoff, color='blue') + 
    geom_text(x=cutoff + .85, y=0.42, label=sprintf('x = %0.2f (%0.3f%%)', cutoff, q))
}
```

```{r qnormplot, exercise=TRUE, exercise.setup='normal_quantile_plot'}
plot_normal_quantile(0.975)
```

```{r qnorm_question}
question("What is the q value that you need to enter in order to find x = 0?",
  answer("0"),
  answer("0.5", correct=TRUE),
  answer("0.95"),
  answer("0.975")
)
```

In particular, note that the values are symmetric -- that is, the value that cuts off the bottom 2.5% is simply the negative version of the value that cuts off the top 2.5%. That's why we can use the $\pm$ in the formula to create the confidence interval.

#### Exercise

Now that you know how to find the appropriate cutoff for the normal distribution, let's create our 95%  confidence interval, using the *my_sample_mean* and *my_sample_sd* variables that you created earlier.You will need to do the following:

- compute the standard error of the mean
- determine the critical value for the normal distribution that cuts off 2.5% of the distribution in each tail, using the `qnorm()` function
- compute the upper and lower limits of the confidence interval by adding or subtracting the appropriate quantity using the formula shown above

```{r ci_exercise, exercise=TRUE, exercise.setup='ci_setup'}
my_sample_SEM <- ...

normal_cutoff <- ...

ci_lower <- ...

ci_upper <- ... 

paste("CI", ci_lower, ci_upper)
```

```{r ci_exercise-solution}
my_sample_SEM <- my_sample_sd / sqrt(sample_size)

normal_cutoff <- qnorm(0.975)

ci_lower <- my_sample_mean - normal_cutoff * my_sample_SEM

ci_upper <- my_sample_mean + normal_cutoff * my_sample_SEM

```

```{r ci_exercise-check}
grade_code(incorrect='Try again...')

```

### Simulating the confidence interval

Here we will simulate the confidence to confirm that it actually works as advertised.  To do this, we will repeatedly sample data from a normal distribution with a true mean of 100 and standard deviation of 10, and compute the 95% confidence interval for the mean. If the confidence interval works correctly, then it should contain the true mean (100) 95% of the time.


```{r simci, exercise=TRUE}
set.seed(12345)
num_simulation_runs <- 5000
sample_size <- 100
ci_percent <- 95
normal_cutoffs_95pct <- c(qnorm((100 - ci_percent)/200),
                          qnorm(1 - (100 - ci_percent)/200))
true_mean = 100
true_sd = 10

captured <- array(dim=num_simulation_runs)

for (sim_run in 1:num_simulation_runs){
  # take the sample
  random_sample <- rnorm(sample_size, mean=true_mean, sd=true_sd)
  # compute the confidence interval
  sample_mean <- mean(random_sample)
  sample_SEM <- sd(random_sample)/sqrt(sample_size)
  sample_CI = sample_mean + sample_SEM*normal_cutoffs_95pct
  captured[sim_run] = sample_CI[1] < true_mean & sample_CI[2] > true_mean
}

mean(captured)
```


This shows us that the confidence interval performs as advertised.  Try running the code chunk above changing the percentage that is specified in the *ci_percent* variable, and confirm that the percentage of captured means 
