---
title: "Vectors"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.eval = FALSE)
knitr::opts_chunk$set(exercise.checker = gradethis::grade_learnr)
```

## Vectors

In this exercise we will go over how to work with vectors.  A vector is a variable that contains more than one value.  For example, let's say you wanted to create a list of states that are not part of the contiguous United States.  You would create that using the ``c()`` function (which stands for "Combine"):

```{r states, exercise=TRUE}
c( 'Alaska', 'Hawaii')
```

A vector can hold any kind of information, as long as all of the entries are of the same type (such as a number or a character string).  Let's say that we want to create a vector containing the population of each state: 731,545 for Alaska, and 1,415,872 for Hawaii.

```{r populations, exercise=TRUE}
c(731415, 1415872)
```

Note that we don't include the commas in numbers, since R uses commas to separate the different elements of the vector, so it would misinterpret them.

Vectors cannot hold values with different modes (types). Let's see what happens if we try to mix the state names (which are character strings) and the population values (which are numeric) within a single vector:

```{r mixed_vector, exercise=TRUE}
c( 'Alaska', 'Hawaii', 731415, 1415872)

```
  
If we look at the result, we see that the numbers are surrounded by quotation marks, which means that R is now treating them as strings rather than numbers.  Later we will see better ways to mix together different types of data within a single variable, using what we call *data frames*.

## Sequences

It's often useful to generate a sequence of numbers.  For example, let's say that you want to generate all of the numbers from 1 to 7, corresponding to the seven days of the week.
You can do this using the following notation:  

*(starting value):(ending value)*

For example:

```{r sequence, exercise=TRUE}
1:7
```


R provides a function called ``seq()`` that is more flexible for generating sequences.
  
```{r seqfunc, exercise=TRUE}
seq(1, 7)

```

One useful feature of ``seq()`` is that we can use arbitrary increments.  Let's say that we wanted to generate values from zero to one, in steps of 0.1  We could do that as follows:

```{r seqincrement, exercise=TRUE}
seq(0, 1, 0.1)

```

We can also use a negative increment to go backwards. Let's say that we want to count how many days are left in the week, from 6 to zero:

```{r seqnegincrement, exercise=TRUE}
seq(6, 0, -1)
```

#### Exercise

Enter the code to generate values starting at five times pi and ending at zero, in steps of pi divided by 2:

```{r seq_exercise, exercise=TRUE}

```

```{r seq_exercise-solution}
seq(5 * pi, 0, -pi/2)
```

```{r seq_exercise-check}
grade_code(incorrect='Try again...')
```


## Accessing the elements of a vector

We often want to access a single element within a vector.  For example, let's say that we have a vector containing an individual's first, middle, and last name:

```{r setup-names, include=FALSE}
names_vector <- c('Russell', 'Allan', 'Poldrack')

```


```{r names, exercise=TRUE}
names_vector <- c('Russell', 'Allan', 'Poldrack')
names_vector
```


Now let's say that we want to extract the middle name from this vector.  We can extract the element in any position by using the position surrounded by square brackets. Since the middle name is in the second position, we would use ``[2]``:  

```{r get_first_name, exercise=TRUE, exercise.setup="setup-names"}
names_vector[2]
```

If you have programmed in other languages, you may know that it's common for other languages to start their indexing at zero rather than one.  It's just one of those features of R that you have to get accustomed to if you have experience in other languages.

We can also use the same indexing to assign new values to elements of a vector.  It turns out that the earlier vector misspelled my middle name -- it should be 'Alan' -- so we can fix it like this:

```{r fixname, exercise=TRUE, exercise.setup="setup-names"}
names_vector[2] <- 'Alan'
names_vector
```

```{r names-fixed, include=FALSE}
names_vector <- c('Russell', 'Alan', 'Poldrack')
```

We can also add a new value to the end of the vector, by using the next index. Let's say we want to add my degree:

```{r addelement, exercise=TRUE, exercise.setup="names-fixed"}
names_vector[4] <- 'Ph.D.'
names_vector
```

```{r names-fixed2, include=FALSE}
names_vector <- c('Russell', 'Alan', 'Poldrack', 'Ph.D.')
```

Sometimes we may want to extract several values from a vector. We can do this by using a another vector as the index.  Let's say that we want to extract just the first and last name from the vector above.  We would pass another vector containing 1 and 3 (the positions of my first and last name):

```{r extract_multiple, exercise=TRUE, exercise.setup='names-fixed2'}
names_vector[c(1, 3)]

```

We can also use a sequence to extract a set of values.  Let's say that we wanted to get my name, leaving out the degree:

This means you can retrieve ranges of values. Get the second through fourth words:

```{r getnames, exercise=TRUE, exercise.setup='names-fixed2'}
names_vector[1:3]
```

#### Exercise

First, create a vector variable called ``first_five_letters`` that contains the first five letters in the alphabet in lower case:

```{r first_five_letters, exercise=TRUE}

```

```{r first_five_letters-solution}
first_five_letters <- c('a', 'b', 'c', 'd', 'e')
```

```{r first_five_letters-check}
grade_code()
```

Now, using a vector index to extract just the vowels ('a' and 'e') and store them to a new variable called ``vowels_only``:

```{r vowels_only, exercise=TRUE, exercise.setup='first_five_letters-solution'}

```

```{r vowels_only-solution}
vowels_only <- first_five_letters[c(1, 5)]
```

```{r vowels_only-check}
grade_code()
```

## Vector Math

You can perform most mathematical operations on vectors just as you would on a variable containing a single value.  

Let's say that we were to measure the high temperature in Palo Alto for five consecutive days, and store it to a variable called ``daily_temperature``:

```{r dailytemp, exercise=TRUE}
daily_temperature <- c(78, 76, 77, 74, 73)
```

Now let's say that we discover a problem with our thermometer, such that it is consistently two degrees higher than the true temperature. Thus, in order to fix our measurements, we need to subtract two from each entry. We can do this by simply subtracting two fromm the vector:

```{r dailytemp-setup}
daily_temperature <- c(78, 76, 77, 74, 73)
daily_temperature_fixed <- daily_temperature - 2
temp_over_72 <- daily_temperature_fixed > 72
daily_temperature_missing <- daily_temperature_fixed
daily_temperature_missing[3] <- NA
```

```{r dailytemp_fix, exercise=TRUE, exercise.setup='dailytemp-setup'}
daily_temperature_fixed <- daily_temperature - 2
daily_temperature_fixed
```

Now say that we want to convert these all to Celsius, using the formula that we used in the earlier example:


$$
T_{Celsius} = (T_{Farenheit} - 32) * \frac{5}{9}
$$

We can simply apply that math to our vector:

```{r convert_temp, exercise=TRUE, exercise.setup='dailytemp-setup'}
daily_temperature_C <- (daily_temperature_fixed - 32) / (5 / 9)
daily_temperature_C
```

## Comparing vectors

We often want to perform a test on each element of a vector.  For example, let's say that we want to know whether the temperature each day is above 72 (which we refer to in Palo Alto as "hot").  We can simply apply our test to the vector, and get back a vector of logical values as the result.

```{r test72, exercise=TRUE, exercise.setup='dailytemp-setup'}
temp_over_72 <- daily_temperature_fixed > 72
temp_over_72
```

#### Exercise

I would like to know how many of the days had a temperature above 72.  Use the ``sum()`` function with the ``temp_over_72`` variable to compute how many days had a temperature over 72, and save this to a new variable called ``days_over_72``. Hint: Remember that R treats ``TRUE`` identically to the number one.


```{r sum_over_72, exercise=TRUE, exercise.setup='dailytemp-setup'}

```

```{r sum_over_72-solution}
days_over_72 <- sum(temp_over_72)
```

```{r sum_over_72-check}
grade_code()
```

This trick will become very important later on when we begin to compute probabilities.

## NA Values 

When we are working with real data, it is common to encounter examples where some of the data are missing. For example, let's say that the thermometer malfunctioned on day three, so that were missing the data for that day.  Rather than eliminating this measurement, we generally want to indicate that it is missing.  R has a built-in value called ``NA`` that denotes that the data are "not available."  Let's make that change to our data:

```{r temp_na, exercise=TRUE,exercise.setup='dailytemp-setup'}
daily_temperature_missing <- daily_temperature_fixed
daily_temperature_missing[3] <- NA
daily_temperature_missing
```

Now let's say that we want to compute the average temperature for those days, which we can compute using the built-in ``mean()`` function:

```{r meanna, exercise=TRUE, exercise.setup='dailytemp-setup'}
mean(daily_temperature_missing)
```

Since one of the values is NA, the mean function doesn't know what to do with it so it returns an NA value.  Many functions have a built-in way to deal with missing values.  Take a look at the help for the ``mean()`` function:

```{r meanhelp, exercise=TRUE}

```

You will see that one of its arguments is ``na.rm`` which tells it to ignore all NA values.

#### Exercise

Compute the mean of ``daily_temperature_missing``, setting the na.rm argument to true:


```{r mean_narm, exercise=TRUE, exercise.setup='dailytemp-setup'}

```

```{r mean_narm-solution}
mean(daily_temperature_missing, na.rm=TRUE)
```

```{r mean_narm-check}
grade_code()
```


