---
title: "Getting started with R"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.eval = FALSE)
knitr::opts_chunk$set(exercise.checker = gradethis::grade_learnr)

```


## Using the R console

Anything that you type into R will be evaluated and the result will be printed to the screen. 

First let's try some simple math. Enter `1 + 1` into the console below, and then click the "Run Code" button, which will execute your command:

```{r one-plus-one, exercise=TRUE}

```

You can see the result (2) printed below the code.

We can also print pieces of text, which are referred to as "strings".  
Type "Hello, world!" in the console below -- be sure to include the quotes!

```{r hello-world, exercise=TRUE}

```

Now try multiplying 5 times 4, using the multiplication operator `*`.


```{r five-times-four, exercise=TRUE}

```

## Logical values

We will often want to determine whether some particular expression is true or false - we refer to these as "logical" values.  

Run the following code, which executes the expression to test whether three is less than four.

```{r three-gt-four, exercise=TRUE}
3 < 4
```

Now enter the expression to test whether 5 is greater than 10:

```{r five-gt-ten, exercise=TRUE}

```

One very common operation is to test whether two numbers are equal.  To do this, we use the double-equals sign (`==`).  

Execute the following code to test whether one plus two is equal to four:

```{r one-plus-two-equals-four, exercise=TRUE}
1 + 2 == 4
```

Now enter the expression to test whether three times three is equal to nine:

```{r three-times-three-equals-nine, exercise=TRUE}

```

One important thing to know is that R treats TRUE the same as 1, and FALSE the same as zero.  This will become very useful when we start working with probabilities.  

Enter an expression to test whether TRUE (capitalization matters here) is equal to 1:

```{r true-equals-one, exercise=TRUE}

```

And do the same to test whether FALSE is equal to zero:

```{r false-equals-zero, exercise=TRUE}

```

## Variables

It is very common in programming to use variables to store particular values.  

For example, let's say that we want to store today's temperature in a variable.  It's always good to name our variables in an explanatory way so that we know what they contain.  There are two ways that we can assign a value to variable in R: using either the ``<-`` operator or the ``=`` operator.  We will use ``<-`` since this is most customary in R.

In this case, let's create a variable called ``todays_temperature_F`` (for today's temperature in Farenheit) and set it to 78:

```{r todays-temp, exercise=TRUE}
todays_temperature_F <- 78
```

You will notice that R does not create any output when you execute this command.

Now try typing the variable name (``todays_temperature_F``) in the console on the line after the variable is defined, and see what the output is:

```{r setup-temp, include=FALSE}
todays_temperature_F <- 78
todays_temperature_C <- (todays_temperature_F - 32) * (5/9)

farenheit_to_celsius = function(temp_F){
  return((temp_F - 32) * (5/9))
}

```

```{r todays-temp-print, exercise=TRUE, exercise.setup = "setup-temp"}

```

Our variable can now be used in place of the original result. Let's create a new variable that converts this value to Celsius, using the following formula:

$$
T_{Celsius} = (T_{Farenheit} - 32) * \frac{5}{9}
$$

```{r todays-temp-convert, exercise=TRUE, exercise.setup = "setup-temp"}
todays_temperature_C = (todays_temperature_F - 32) * (5/9)
todays_temperature_C
```

#### Exercise

Enter the code to test whether today's temperature in Celsius is above freezing (which is zero Celsius):

```{r todays-temp-above-freezing, exercise=TRUE, exercise.setup = "setup-temp"}

```

```{r todays-temp-above-freezing-solution}
todays_temperature_C > 0
```

```{r todays-temp-above-freezing-check}
grade_code(incorrect='Try again...')
```

## Using functions

A function in R is similar to a function in mathematics: It takes in an input, and returns an output.  For example, we might have a function (let's call it ``farenheit_to_celsius()``) that would take in a temperature in Farenheit and return the value converted to Celsius.

If we want to apply this function to our temperature, we simply call the function with our variable as the *argument* to the function, which is contained in the parentheses after the function name:

```{r apply_f_to_c, exercise=TRUE, exercise.setup = "setup-temp"}
farenheit_to_celsius(todays_temperature_F)
```

Because we didn't assign the output from the function to a new variable, R prints the value to the console.

Many functions can accept multiple values. For example, let's say that we want to sum of three numbers:

```{r take_sum, exercise=TRUE}
sum(4, 7, 9)
```

To find out more about a function, you can use the question mark sign to get help.  For example, let's say that I have asked you to use the ``rep()`` function to create a set of repeated numbers.  To learn more about that function, you could do the following:

```{r get_help_rep, exercise=TRUE}
?rep
```

Some functions, including ``rep()``, have additional arguments that are named.  For example, if you want to rep to repeat the number 5 seven times, you would set the "times" argument to 7:

```{r rep_5_7, exercise=TRUE}

rep(5, times=7)
```

#### Exercise

In the following console, enter the code that would be necessary to compute the cosine of 2 times pi.  Note that ``pi`` is a built-in variable in R that contains the value for pi, so you don't need to type in the actual number.  The cosine is computed using the ``cos()`` function.  Click "Submit Answer" to check whether your answer is correct.

```{r cos_2_pi, exercise=TRUE}

```

```{r cos_2_pi-check}
grade_result(
  pass_if(~ identical(.result, cos(2*pi)), "Good job!"),  
  fail_if(~ TRUE, "Not quite right, try again!.")

)
```

